Predicting Fire Risk: \\ An Evaluation of the Fire Danger Index

### 2016-01-24 paper.tex
Improved latex formatting

\subsection{Predicting fire activity}
A study linking fire activity to certain climate variables (some of which are inputs to the FFDI, others of which are not) and the El Niño Southern Oscillation (ENSO) index found significant relationships between fire activity and vapour pressure, maximum and minimum temperature and antecedent rainfall \citep{harris14}. Notably the variables which the indices are most sensitive to are different to those found to have the strongest link with fire activity suggesting that the fire danger indices may not contain all the important information for predicting fire activity \citep{harris14}.

@@

The meteorological data set comes from a climate simulation of the Murray Darling Basin using the Weather Research and Forecasting (WFR) model. The WRF Modeling system has been developed by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administrations Centers for Environmental Prediction (NCEP) and the other weather research institutions in the United States. \citet{evans10} used the Advanced Research WFR version 3 \citep{skamarock08} to simulate weather for southeastern Australia at a 10km horizontal grid over 24 years from 1985.

@@

\subsubsection{Grassland Curing}
The level of dying or senescence of plant material caused by seasonal weather patterns is referred to as grassland curing. Grassland curing ranges between 0 and 100, where 0 means the grass is green with high moisture content and 100 means the grass is at its driest. Curing has a significant impact on fire behaviour in grasslands, particularly the probability of ignition and subsequent rate of spread \citep{cheney08}.

@@

\subsubsection{Vapour Pressure}
Vapour pressure, in meteorology, describes the partial pressure of water vapour in the atmosphere, it is directly related to the number of water vapour molecules in the air.  \citet{harris14} find there is a significant relationship between fire activity and vapour pressure among other climate variables. I investigate whether vapour pressure at 9am can improve prediction of fire occurrence.

@@

%eg confusion matrix with ABCD
\begin{table}
	\centering
	\begin{tabular}{lrrrr}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
		\cline{3-5}
\emph{Actual} & No Fire & $a$   & $b$   & $a+b$ \\
                 & 		Fire & $c$   & $d$   & $c + d$ \\
		\cline{3-5}
                 & 		     & $a+c$ & $b+d$ & $a+b+c+d$ \\
		\cline{3-5}
	\end{tabular}
	\caption{Example confusion matrix}
	\label{table:ex}
\end{table}
An example confusion matrix is given in Table~\ref{table:ex}. 
From this matrix, we can compute accuracy, sensitivity (correctly classified fires dividied by total number of actual fires), specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
\label{eq acc}
\mbox{Accuracy} &= \frac{a + d}{a+b+c+d} \\
\label{eq sens}
\mbox{Sensitivity} &= \frac{a }{a+b} \\
\label{eq spec}
\mbox{Specificity} &= \frac{d}{c+d} \\
\label{eq BA}
\mbox{Balanced Accuracy} &= \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
\label{eq:K}
\kappa &= \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

@@

%Output from best performing splines model (used xtable rather than texreg, too inconsistent?)
\begin{table}
	\centering
	\begin{tabular}{rrrrr}
		\toprule
		& Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
		\midrule
		(Intercept) & -3.1484 & 0.2757 & -11.4210 & 0.0000 \\ 
		ns(ffdi, 5)1 & -0.2905 & 0.3613 & -0.8043 & 0.4213 \\ 
		ns(ffdi, 5)2 & -0.2304 & 0.4625 & -0.4982 & 0.6183 \\ 
		ns(ffdi, 5)3 & 0.4040 & 0.5589 & 0.7229 & 0.4697 \\ 
		ns(ffdi, 5)4 & -0.4340 & 0.9674 & -0.4487 & 0.6537 \\ 
		ns(ffdi, 5)5 & -0.8815 & 0.7271 & -1.2124 & 0.2254 \\ 
		ns(temp, 1) & 1.8543 & 0.3416 & 5.4276 & 0.0000 \\ 
		ns(rh, 2)1 & 0.9325 & 0.5687 & 1.6397 & 0.1011 \\ 
		ns(rh, 2)2 & -0.6173 & 0.3607 & -1.7115 & 0.0870 \\ 
		ws & -0.0169 & 0.0146 & -1.1541 & 0.2485 \\ 
		ns(df, 5)1 & 1.1093 & 0.2798 & 3.9639 & 0.0001 \\ 
		ns(df, 5)2 & 1.0156 & 0.3115 & 3.2601 & 0.0011 \\ 
		ns(df, 5)3 & 0.7113 & 0.2604 & 2.7319 & 0.0063 \\ 
		ns(df, 5)4 & 0.8071 & 0.5479 & 1.4730 & 0.1408 \\ 
		ns(df, 5)5 & -0.1512 & 0.1950 & -0.7752 & 0.4382 \\ 
		ns(curing, 6)1 & 1.5648 & 0.1388 & 11.2754 & 0.0000 \\ 
		ns(curing, 6)2 & 2.3986 & 0.1874 & 12.7997 & 0.0000 \\ 
		ns(curing, 6)3 & 1.6637 & 0.1640 & 10.1436 & 0.0000 \\ 
		ns(curing, 6)4 & 1.6949 & 0.1335 & 12.6921 & 0.0000 \\ 
		ns(curing, 6)5 & 1.4634 & 0.3436 & 4.2588 & 0.0000 \\ 
		ns(curing, 6)6 & -0.4909 & 0.0817 & -6.0066 & 0.0000 \\ 
		rain & -0.0286 & 0.0056 & -5.0703 & 0.0000 \\ 
		vap & -0.0003 & 0.0067 & -0.0504 & 0.9598 \\ 
		\bottomrule
	\end{tabular}
	\caption{Regression output for Model 4}
	\label{table:logf}
\end{table}

@@

Figure~\ref{fig:ffdi} is the ROC curve for the four logistic models. The ROC curve for model 1 shows the FFDI is barely better than 'random assignment' (given by 45 degree line) while inclusion of the FFDI inputs and three additional variables improves the ROC curve, clearly the superior model is Model 4. Allowing for variables to have a non-linear effect on the log-odds ratio had improved the prediction accuracy significantly.

@@

  \begin{tabular}{rcrrcrr}
        & \multicolumn{2}{c}{Tree 1} & 
        & \multicolumn{2}{c}{Tree 2} & \\
    \cmidrule{2-3}\cmidrule{5-6}
        & \multicolumn{2}{c}{Actual} & 
        & \multicolumn{2}{c}{Actual} &  \\
		Predict  & No Fire & Fire && No Fire & Fire\\ 
    \cmidrule{2-3}\cmidrule{5-6}
No Fire  & 0.5553 & 0.0017 && 0.5088 & 0.0014 \\ 
Fire  & 0.4394 & 0.0036 && 0.4859 & 0.0038 \\ 
    \cmidrule{2-3}\cmidrule{5-6}
	\end{tabular}
	\caption{Confusion Matrices for trees cutoff value 0.5}
	\label{table:tcm2}
\end{table}

>>>>

\subsection{Predicting fire activity}

A study linking fire activity to certain climate variables (some of which are inputs to the FFDI, others of which are not) and the El Niño Southern Oscillation (ENSO) index found significant relationships between fire activity and vapour pressure, maximum and minimum temperature and antecedent rainfall \citep{harris14}. Notably the variables which the indices are most sensitive to are different to those found to have the strongest link with fire activity suggesting that the fire danger indices may not contain all the important information for predicting fire activity \citep{harris14}.

@@

The meteorological data set comes from a climate simulation of the Murray Darling Basin using the Weather Research and Forecasting (WFR) model. The WRF Modeling system has been developed by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administrations Centers for Environmental Prediction (NCEP) and the other weather research institutions in the United States. \citet{evans10} used the Advanced Research WFR version~3 \citep{skamarock08} to simulate weather for southeastern Australia at a 10km horizontal grid over 24 years from 1985.

@@

\subsubsection{Grassland Curing}

The level of dying or senescence of plant material caused by seasonal weather patterns is referred to as grassland curing. Grassland curing ranges between 0 and 100, where 0 means the grass is green with high moisture content and 100 means the grass is at its driest. Curing has a significant impact on fire behaviour in grasslands, particularly the probability of ignition and subsequent rate of spread \citep{cheney08}.

@@

\subsubsection{Vapour Pressure}

Vapour pressure, in meteorology, describes the partial pressure of water vapour in the atmosphere, it is directly related to the number of water vapour molecules in the air.  \citet{harris14} find there is a significant relationship between fire activity and vapour pressure among other climate variables. I investigate whether vapour pressure at 9am can improve prediction of fire occurrence.

@@

%eg confusion matrix with ABCD
\begin{table}
  \centering
  \begin{tabular}{lrrrr}
    & 	\multicolumn{1}{l}{} & \multicolumn{2}{c}{\emph{Model Prediction}} \\
    & 	\multicolumn{1}{l}{} & \multicolumn{1}{r}{No Fire} & \multicolumn{1}{r}{Fire} \\
    \cline{3-5}
    \emph{Actual} & No Fire & $a$   & $b$   & $a+b$ \\
                  &   Fire  & $c$   & $d$   & $c + d$ \\
    \cline{3-5}
                  &         & $a+c$ & $b+d$ & $a+b+c+d$ \\
    \cline{3-5}
  \end{tabular}
  \caption{Example confusion matrix}
  \label{table:ex}
\end{table}
An example confusion matrix is given in Table~\ref{table:ex}.
From this matrix, we can compute accuracy, sensitivity (correctly classified fires dividied by total number of actual fires), specificity (correctly identified non-fires divided by total number of non-fires), balanced accuracy and kappa:
\begin{align}
  \label{eq acc}
  \mbox{Accuracy}          & = \frac{a + d}{a+b+c+d}                     \\
  \label{eq sens}
  \mbox{Sensitivity}       & = \frac{a }{a+b}                            \\
  \label{eq spec}
  \mbox{Specificity}       & = \frac{d}{c+d}                             \\
  \label{eq BA}
  \mbox{Balanced Accuracy} & = \frac{\frac{a }{a+b} + \frac{d}{c+d} }{2} \\
  \label{eq:K}
  \kappa                   & = \frac{P_0 - P_c}{1-P_c}
\end{align}
where $P_0$ is the total agreement probability and $P_c$ is the Agreement probability which is due to chance (see \cite{kappa} for definitions).

@@

%Output from best performing splines model (used xtable rather than texreg, too inconsistent?)
\begin{table}
  \centering
  \begin{tabular}{rrrrr}
    \toprule
                   & Estimate & Std. Error & z value  & Pr($>|z|$) \\
    \midrule
    (Intercept)    & -3.1484  & 0.2757     & -11.4210 & 0.0000 \\
    ns(ffdi, 5)1   & -0.2905  & 0.3613     & -0.8043  & 0.4213 \\
    ns(ffdi, 5)2   & -0.2304  & 0.4625     & -0.4982  & 0.6183 \\
    ns(ffdi, 5)3   & 0.4040   & 0.5589     & 0.7229   & 0.4697 \\
    ns(ffdi, 5)4   & -0.4340  & 0.9674     & -0.4487  & 0.6537 \\
    ns(ffdi, 5)5   & -0.8815  & 0.7271     & -1.2124  & 0.2254 \\
    ns(temp, 1)    & 1.8543   & 0.3416     & 5.4276   & 0.0000 \\
    ns(rh, 2)1     & 0.9325   & 0.5687     & 1.6397   & 0.1011 \\
    ns(rh, 2)2     & -0.6173  & 0.3607     & -1.7115  & 0.0870 \\
    ws             & -0.0169  & 0.0146     & -1.1541  & 0.2485 \\
    ns(df, 5)1     & 1.1093   & 0.2798     & 3.9639   & 0.0001 \\
    ns(df, 5)2     & 1.0156   & 0.3115     & 3.2601   & 0.0011 \\
    ns(df, 5)3     & 0.7113   & 0.2604     & 2.7319   & 0.0063 \\
    ns(df, 5)4     & 0.8071   & 0.5479     & 1.4730   & 0.1408 \\
    ns(df, 5)5     & -0.1512  & 0.1950     & -0.7752  & 0.4382 \\
    ns(curing, 6)1 & 1.5648   & 0.1388     & 11.2754  & 0.0000 \\
    ns(curing, 6)2 & 2.3986   & 0.1874     & 12.7997  & 0.0000 \\
    ns(curing, 6)3 & 1.6637   & 0.1640     & 10.1436  & 0.0000 \\
    ns(curing, 6)4 & 1.6949   & 0.1335     & 12.6921  & 0.0000 \\
    ns(curing, 6)5 & 1.4634   & 0.3436     & 4.2588   & 0.0000 \\
    ns(curing, 6)6 & -0.4909  & 0.0817     & -6.0066  & 0.0000 \\
    rain           & -0.0286  & 0.0056     & -5.0703  & 0.0000 \\
    vap            & -0.0003  & 0.0067     & -0.0504  & 0.9598 \\
    \bottomrule
  \end{tabular}
  \caption{Regression output for Model 4}
  \label{table:logf}
\end{table}

@@

Figure~\ref{fig:ffdi} is the ROC curve for the four logistic models. The ROC curve for model 1 shows the FFDI is barely better than `random assignment' (given by 45 degree line) while inclusion of the FFDI inputs and three additional variables improves the ROC curve, clearly the superior model is Model 4. Allowing for variables to have a non-linear effect on the log-odds ratio had improved the prediction accuracy significantly.

@@

  \begin{tabular}{rcrrcrr}
    & \multicolumn{2}{c}{Tree 1} &
    & \multicolumn{2}{c}{Tree 2} & \\
    \cmidrule{2-3}\cmidrule{5-6}
    & \multicolumn{2}{c}{Actual} &
    & \multicolumn{2}{c}{Actual} &  \\
    Predict & No Fire & Fire   & & No Fire & Fire\\
    \cmidrule{2-3}\cmidrule{5-6}
    No Fire & 0.5553  & 0.0017 & & 0.5088  & 0.0014 \\
    Fire    & 0.4394  & 0.0036 & & 0.4859  & 0.0038 \\
    \cmidrule{2-3}\cmidrule{5-6}
  \end{tabular}
  \caption{Confusion Matrices for trees cutoff value 0.5}
  \label{table:tcm2}
\end{table}
